Word-Referent Identification Under Multimodal Uncertainty

### 2018-02-24 Writeup/ms.Rmd
second draft journal write-up

---
title             : "Word Identification Under Multimodal Uncertainty"
shorttitle        : "Word Identification Under Multimodal Uncertainty"

@@

  Deprtment of Psychology

  Stanford Univerisity

@@

abstract: |
 Identifying the visual referent of a spoken word -- that a particular insect is referred to by the word ``bee'' -- requires both the ability to process and integrate multimodal input and the ability to reason under uncertainty. How do these tasks interact with one another? We introduce a task that allows us to examine how adults identify words under joint uncertainty in the auditory and visual modalities. We propose an ideal observer model of the task which provides an account of how auditory and visual cues are combined optimally. Model predictions are tested in three experiments where word recognition is made under two kinds of uncertainty: category ambiguity and/or distorting noise. In all cases, the optimal model explains much of the variance in human judgments. In particular, when the signal is not distorted with noise, participants weight the auditory and visual cues optimally, that is, according to the relative reliability of each modality. But when one modality has noise added to it, human perceivers systematically prefer the unperturbed modality to a greater extent than the optimal model does. We discuss the broad implications of the results, especially in relation to early language acuquisition.

keywords          : "Language; audio-visual processing; word learning; speech perception; computational modeling."

@@

#Data from the 3 experiments 
#############################
exp1 <- read_delim("../Analysis/exp1_data_NoAnonym.txt", delim = " ") %>%
  filter(type == "Task") %>%
  mutate(experiment='Experiment1')

exp2 <- read_delim("../Analysis/exp2_data_NoAnonym.txt", delim = " ") %>%
  filter(type == "Task") %>%
  mutate(experiment='Experiment2')

exp3 <- read_delim("../Analysis/exp3_data_NoAnonym.txt", delim = " ") %>%
  filter(type == "Task") %>%
  mutate(experiment='Experiment3')
```

@@

In this line of research, @clayard08 simulated auditory uncertainty by manipulating the probability distribution of a cue (VOT) that differentiated similar words (e.g., "beach" and "peach"). They found that humans were sensitive to these probabilistic cues and their judgments closely reflected the optimal predictions. In another work, @feldman2009 studied the perceptual magnet effect, that is, the fact that categories influence the way the stimuli are perceived. They showed that this effect can be explained as the consequence of optimally solving the problem of perception under uncertainty. Both studies explored optimal performance under uncertainty from one modality, i.e., the auditory modality. There is however extensive evidence that information from the visual modality, in particular the speaker's facial features,  influences the perception of auditory information [see @Campbell2008 for a review]. @bejjanki2011 offered a mathematical characterization of how cues from speech and lip movements can be optimally combined and showed that human performance during audio-visual phonemic labeling was consistent (at least at the qualitative level) with the behavior of the ideal observer.

The speaker's facial features are not the only visual information that can be used to disambiguate the auditory input, however. For instance, in a referential context, information about the identity of the semantic referent can be integrated with linguistic information to resolve lexical and syntactic ambiguities in speech [e.g., @Eberhard1995; @Tanenhaus1995; @spivey2002]. To our knowledge, however, no study offered an ideal observer analysis of word identification in such context, that is, when the mind has to combine cues from the sound and the referent. Imagine, for example, that someone is uncertain whether they heard "pea" or "bee", does this uncertainty make them rely more on the referent (e.g., the object being pointed at)? Vice versa, if they are not sure if they saw a bee or a fly, does that make them rely more on the sound? More importantly, when input in both modalities is uncertain to varying degrees, do they weight each modality according to its relative reliability (which is the optimal strategy), or do they over-rely on a particular modality (which is a sub-optimal strategy)?

@@

First, in the case of audio-visual speech, modalities are understood to differ only in terms of their reliability. In a referential context, however, modalities also differ in terms of their roles in the referential process: the auditory input represents the symbol (i.e., the word) whereas the visual input represents the referent. It has been suggested that because of the referential property of speech, it is a privileged signal for humans starting in infancy [see @vouloumanos2014 for a review].\footnote{There is a debate over whether speech is privileged for infant and adults for the same reasons. Whereas Waxman and colleagues suggest that speech is privileged for both infants and adults because of its ability to refer, Sloutsky and colleagues suggest that speech might not have a referential status from the start. Rather, they argue tha speech seems to be prefered by infants only because of a low level auditory "overshadowing".} Thus, in a referential context, it is possible that we do not treat the auditory and visual modalities as equivalent sources of information. Instead, there could be a bias for the auditory modality beyond what is expected from informational reliability alone.

@@

We test the ideal observer model's predictions in three behavioral experiments where we vary the source of uncertainty. In Experiment 1, audio-visual tokens are ambiguous with respect to their category membership only. In Experiment 2, we intervene by adding background noise to the auditory modality, and in Experiment 3, we intervene by adding background noise to the visual modality. In all experiments, participants were quantitatively near-optimal, though overall response precision was slightly lower than expected. Moreover, in Experiment 1 where neither of the modalities was perturbed with background noise, participants weighed auditory and visual cues according to the relative reliability predicted by the optimal model. In other words, we found no evidence for a modality bias towards either the auditory or the visual modality. However, in Experiment 2 and 3, participants over-relied on one modality when the other modality was perturbed with additional noise. <!-- In the discussion, we talk about the broad implications of these results and especially in relation to some developmental findings in early word learning [e.g., @stager1997]. -->

@@

In this section we describe the multimodal integration model. First, we briefly introduce the experimental paradigm, and second we explain how behavior in this paradigm can be characteized optimally with an ideal observer model.

@@

```{r task, fig.cap = "Overview of the task. In the audio-visual condition, participants are first presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test. The test may differ from the target in both the auditory and the visual components. After these two presentations, participants press same' (i.e., the same category as the target) or `different' (not the same category). The auditory-only and visual-only conditions are similar, expect only the sounds are heard, or only the pictures are shown, respectively.", fig.align = "center", out.width = "400px"}
knitr::include_graphics("pictures/task.png", dpi = 108)
```

This task is similar to the task introduced by @sloutsky2003, and used in subsequent studies in both children and adults [see @robinson2010 for a review]. However, in this original task, participants are asked whether or not the two audio-visual presentations are *identical*. Such task would allow us to probe audio-visual encoding, but not necessarily language understanding. The latter requires---in addition to the perceptual encoding---a categorical judgement, i.e, determining wheather or not two similar tokens are members of the same phonological/semantic category. Thus, our task is category-based: participants are asked to press "same" if they think the second item (the test) belonged to the same category as the first (target) (e.g.,  dog-/aba/), even if there is a slight difference in the sound, in the referent, or in both. They are instructed to press "different" only if they think that the second stimulus was an instance of the other word category (cat-/ada/). The task also includes trials where pictures are hidden (audio-only) or where sounds are muted (visual-only). These unimodal trials provide us with participants' evaluation of the probabilistic information present in the auditory and visual categories. This unimodal distributions are used as inputs to the optimal combination model, described below.

@@

The ideal observer model has to combine probabilistic information from auditory and the visual modalities and make a categorical decision. How can this task be characterized in a formal way, and what would be the optimal behavior? If the stimuli were continuous (e.g., size, location), cues derived from each modality would be characterized with their sensory noise variance only, and (assuming conditional independence) their combined information would be characterized with a simple linear combination rule where the combined estimate is an average of the indivisual estimates, each wighted according to its relative reliability [@ernst02; @alais04]. However, in our case decison is made over categorical variables (phonemes and concepts), the optimal model should take into account, not only the noise variability arround an individual perceptual estimate, but also its categorical variability (i.e., the probability with which the estimate belongs to a given category).

One way to account for the categorical variability is to model the decision step which maps sensory input onto task-dependent categories [@Bankieris17; @bejjanki2011]. We represent the decision step as follows (see Figure\ \@ref(fig:model) for illustration). If we assume that both the visual and the auditory cues vary along a single dimension, an audio-visual signal $(a,v)$ can be represented as a point in this two-dimensionl space, and a word category can be characterized with a bivariate probability distribution over audio-visual tokens.\footnote{For the sake of clarity, we assume here that the sensory noise around this point is small/negligeable compared to the categorical vriability, but see below how the noise can be integrated when its effect is realtively large} Each category is peaked around the most typical audio-visual token, and the probability of membreship in the category (the likelihood probability) decreases gradually as the auditory token deviates from the typical sound instance (depending on the variance of the auditory category) and/or as the visual token deviates from the typical visual instance (depending on the variance of the visual category). <!--To illustrate how categorical decision is made within this formal framework, suppose we receive an ambiguous image which can be either of a cat or of a dog, associated with a slightly noisy auditory input that sounds more like "cat". If the judgment is based on the picture only, this would lead to a random choice between cat and dog. However, if the judgement is based on both the picture and the sound (which is less ambiguous), the combined information should lead to a preference for the word cat  we illustrate the cateorical decision making. Audio-visual points along this line lead to random responses, meaning that both the auditory and the visual components are maximally ambiguous. Audio-visual points that lay on either side of the decision threshold lead to decisions in favor of one word or the other. This means that at least one component of this audio-visual token is not totally ambiguous, which leads to a diambiguated combined information. -->
A crucial feature of the ideal observer model is that it makes use of the entire probability distribution of a given category in order to make optimal decisions [e.g., @clayard08]. This means that the probability of picking a word (the posterior probability) should mirror closely the probability of the combined information (the likelihood probability). For instance, the model becomes less ambiguous and more certain as the tokens get closer to the typical instance of the category. This intuition will be formalized below.
<!--characterized with a likelihood function, e.g., $P(S_V | S)$ and $P(S_A | S)$ for a visual and an auditory cue, respectively. If these cues are conditionally independent, their combined information can be written as $P(S_V, S_A | S)=P(S_A | S)P(S_V | S)$. The maximum likelihood estimate of this combined information $\hat{S}$ is an average of the individual maximum liklihoods $\hat{S}=w_V\hat{S_V}+w_A\hat{S_A}$, each wighted according to its relative reliability:
$w_V=\frac{1/\sigma_V^2}{1/\sigma_V^2 + 1/\sigma_A^2}$, and $w_A=\frac{1/\sigma_A^2}{1/\sigma_V^2 + 1/\sigma_A^2}$.

@@

An auditory categoy $A$ and a visual category $V$ are defined as distributions in the auditory and the visual dimensions, respectively. A word category $W$ is defined as a bivariate distribution in the audio-visual. More specifically, we define it as the joint distribution of the auditory and the visual categories.  For simplicity, we assume both auditory and visual categories to be normally distributed. That is, if $w=(a,v)$ is an audio-visual token, then:
$$ p(a | A) \sim  N(\mu_A, \sigma^2_A) $$
where $\mu_A$ and $\sigma^2_A$ are respectively the mean and the variance of the auditory category,
$$ p(v | V) \sim  N(\mu_V, \sigma^2_V) $$
where $\mu_V$ and $\sigma^2_V$ are  the mean and the variance of the visual category, and

@@

This assumption simply says that, given a word-object mapping, e.g., W=("cat"-CAT), variation in the way "cat" is pronounced does not correlate with changes in any visual property of the object CAT, which is a valid assumption.\footnote{Note that this assumptions is more adequate to the case of arbitrary associations such as ours, and less so in the case of redundant association such as audio-visual speech. In the latter, variation in the pronounciation is expected to correlate, at least to some extent, with lip movements.}

Now we turn to the crucial question of modeling how optimal decision should proceed based on the probabilistic information in the auditory and the visual modalities. In our task, we have two word categories: dog-/aba/ ($W_1$) and cat-/ada/ ($W_2$).\footnote{This mapping is randomized in the experiments.} Participants can be understood as choosing one of these two word categories (Figure\ \@ref(fig:model)). For an ideal observer, the probability of choosing category 2 when presented with an audio-visual instance $w=(a,v)$ is the posterior probability of this category:
$$
p(W_2 | w)=\frac{p(w|W_2)p(W_2)}{p(w|W_2)p(W_2)+p(w|W_1)p(W_1)}
$$
Using our assumption that the cues are uncorrelated, we have
$$p(w | W) = p(a,v| W) = p(a| A)p(v| V)$$
Under this assumption, the posterior probability reduces to:\footnote{see the appendix for the details of the derivation}
\begin{equation}
 p(W_2 | w)=\frac{1}{1+(1+\epsilon)\exp(\beta_0+\beta_aa+\beta_vv)}
\end{equation}
where 
$$1+\epsilon=\frac{p(W_1)}{p(W_2)}$$
$$\beta_0=\frac{\mu^2_{A2}-\mu^2_{A1}}{2\sigma^2_{A}}+\frac{\mu^2_{V2}-\mu^2_{V1}}{2\sigma^2_{V}}$$

@@

The parameter $\epsilon$ represents the differential between the categories' prior probabilities. However, since the identity of word categories is randomized across participants, $\epsilon$ measures, rather, a response bias to "same" if $\epsilon > 0$, and a response bias to "different" if $\epsilon < 0$. We expect a general bias towards answering "different" because of the categorical nature of our same-different task: when two items are ambiguous but perceptually different, participants might have a slight preference for "different" over "same".

@@

## Auditory and Visual baseline models

Besides the optimal model, we also test the predictions of two baselines models: a visual model, which assumes that participants rely only on visual information, and a auditory model, which assumes that participants rely only on auditory information. If participants rely on both the auditory and the visual modalities, the optimal model would explain more variance in human responses than the visual model alone or the auditory model alone.

@@

A systematic comparison of the optimal and the descriptive models would allow us, not only to quantify how much people deviate from optimality, but also to understand precisely how they deviate from this optimality. Let $\sigma^2_{A}$ and $\sigma^2_{V}$ be the values of the variances used in the optimal model (derived from the unimodal conditions), and $\sigma^2_{Ab}$ and $\sigma^2_{Vb}$ be the values observed through the descriptive model in the bimodal condition. Deviation from optimality can be quantified in two ways. First, we quantify changes in the values of the variances specific to each modality, that is, how $\sigma^2_{A}$ compares to $\sigma^2_{Ab}$, and how $\sigma^2_{V}$ compares to $\sigma^2_{Vb}$. Second,  we quantify changes in the proportion of the visual and auditory variances, i.e., how $\frac{\sigma^2_{A}}{\sigma^2_{V}}$ compares to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}$. The first comparison allows us to test if precision varies for each modality, and the second, allows us to test whether or not the weighing scheme folows the prediction of the optimal model.  The reason we used the proportion of the variances as an measure of cross-modal weighing is because it correponds to the slope of the decision threshold in the audio-visual space (Figure\ \@ref(fig:model)). The decision threshold is defined as the set of values in this audio-visual space along which the posterior is equal to 0.5. Formally speaking, the decision threshold takes the following form:

@@

If the slope derived from the descriptive model is greater than the slope of the optimal model, the corresponding shift in the decision threshold indicates that participants have a preference for the auditory modality in the bimodal case. Similarly, a smaller slope would lead to a preference for the visual modality. The limit cases are when there is exclusive reliance on the auditory cue (a vertical line), and where there is exclusive reliance on the visual (a horizontal line).

There are three possible scenarios for sub-optimality, as illustrated in Figure\ \@ref(fig:subOptim):

1) Both variances may increase, but their proportion remains the same. That is, $\frac{1}{\sigma^2_{Ab}} \leqslant \frac{1}{\sigma^2_{A}}$ and $\frac{1}{\sigma^2_{Vb}} \leqslant \frac{1}{\sigma^2_{V}}$, but  $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}} \approx \frac{\sigma^2_{A}}{\sigma^2_{V}}$. In this case, sub-optimality would be due to increased randomness in human responses in the bimodal condition. However, this randomness would not affect the relative weighting of both modalities, i.e., participants would still weigh modalities according to the relative reliability predicted by the optimal model.

2) The auditory variance increases at a higher rate.  That is, $\frac{1}{\sigma^2_{Ab}} \ll \frac{1}{\sigma^2_{A}}$ and $\frac{1}{\sigma^2_{Vb}} \leq \frac{1}{\sigma^2_{V}}$, leading to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}} > \frac{\sigma^2_{A}}{\sigma^2_{V}}$. In this case, sub-optimally would consist not only in participants being more random in the bimodal condition, but also in having a systematic preference for the visual modality, even after accounting for informational reliability.

3) The visual variance increases at a higher rate. That is, $\frac{1}{\sigma^2_{Vb}} \ll \frac{1}{\sigma^2_{V}}$, and  $\frac{1}{\sigma^2_{Ab}} \leq \frac{1}{\sigma^2_{A}}$, leading to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}} > \frac{\sigma^2_{A}}{\sigma^2_{V}}$. This case is the reverse of case 2, i.e., in addition to increased randomness in the bimodal condition, there is a systematic preference for the auditory modality, even after accounting for informational reliability.

```{r subOptim, fig.cap = "Illustration using simulated data showing the prediction of the optimal model (top), and the three possible ways human responses can deviate from the optimal predictions (bottom). These cases are: 1) the variance increases equally for both modalities, but the cue weighting is still optimal, 2) The auditory variance increases at a higher rate, leading to a preference for the auditory modality, and 3) The visual variance increases at a higher rate, leading to a preference for the visual modality", fig.align = "center", out.width = "400px"}
knitr::include_graphics("pictures/sub-optimal", dpi = 108)
```

We tested these predictions in three experiments. In Experiment 1, we studied the case where bimodal uncertainty is due to ambiguity in terms of category membership, without any additional background noise. In Experiment 2 and 3 we added background noise on top of ambiguity in category membership.

@@

In this Experiment, we start with testing the predictions in the case where uncertainty is due to ambiguity in terms of category membership only, that is, according to the following weighting scheme:  $$\beta_a \propto \frac{1}{\sigma^2_{A}}$$ $$\beta_v \propto \frac{1}{\sigma^2_{V}}$$ We do not introduce any additional noise in the background, and we assume that sensory noise is negligeable compared to categorical variability.

@@

```{r bimodal, fig.cap = "Human responses vs. models' predictions across the three Experiments. Shape represents auditory distance from the target, and color represents visual distance from the target.", fig.align = "center", out.width = "400px", fig.width=7, fig.height=8}

>>>>

---
title             : "Word-Referent Identification Under Multimodal Uncertainty"
shorttitle        : "Word Identification Under Multimodal Uncertainty"

@@

  Department of Psychology

  Stanford University

@@

abstract: |

 Identifying a spoken word in a referential context requires both the ability to process and integrate multimodal input and the ability to reason under uncertainty. How do these tasks interact with one another? We introduce a task that allows us to examine how adults identify words under joint uncertainty in the auditory and visual modalities. We propose an ideal observer model of the task which provides an account of how auditory and visual cues are combined optimally. Model predictions are tested in three experiments where word recognition is made under two kinds of uncertainty: category ambiguity and/or distorting noise. In all cases, the optimal model explains much of the variance in human mean judgments. In particular, when the signal is not distorted with noise, participants weight the auditory and visual cues optimally, that is, according to the relative reliability of each modality. But when one modality has noise added to it, human perceivers systematically prefer the unperturbed modality to a greater extent than the optimal model does. The study provides a formal framework which helps us understand precisely how word form and word meaning interact in word recognition under uncertainty. Moreover it offers a first step towards a model that accounts for form-meaning synergy in early word learning.

keywords          : "Language understanding; audio-visual processing; word learning; speech perception; computational modeling."

@@

#Data from the 3 experiments 
#############################
exp1 <- read_delim("../Data_and_analysis/data_exp1_anonym.txt", delim = " ") %>%
  filter(type == "Task") %>%
  mutate(experiment='Experiment1')

exp2 <- read_delim("../Data_and_analysis/data_exp2_anonym.txt", delim = " ") %>%
  filter(type == "Task") %>%
  mutate(experiment='Experiment2')

exp3 <- read_delim("../Data_and_analysis/data_exp3_anonym.txt", delim = " ") %>%
  filter(type == "Task") %>%
  mutate(experiment='Experiment3')
```

@@

In this line of research, @clayard08 simulated auditory uncertainty by manipulating the probability distribution of a cue (VOT) that differentiated similar words (e.g., "beach" and "peach"). They found that humans were sensitive to these probabilistic cues and their judgments closely reflected the optimal predictions. In another work, @feldman2009 studied the perceptual magnet effect, which is related to how categories influence the way the stimuli are perceived. They showed that this effect can be explained as the consequence of optimally solving the problem of perception under uncertainty. Both studies explored optimal performance under uncertainty from one modality, i.e., the auditory modality. There is however extensive evidence that information from the visual modality, in particular the speaker's facial features,  influences the perception of auditory information [see @Campbell2008 for a review]. @bejjanki2011 offered a mathematical characterization of how cues from speech and lip movements can be optimally combined and showed that human performance during audio-visual phonemic labeling was consistent (at least at the qualitative level) with the behavior of the ideal observer.

The speaker's facial features are not the only visual information that can be used to disambiguate the auditory input, however. For instance, in a referential context, information about the identity of the semantic referent can be integrated with linguistic information to resolve lexical and syntactic ambiguities in speech [e.g., @Eberhard1995; @Tanenhaus1995; @spivey2002]. To our knowledge, however, no study offered an ideal observer analysis of word identification in such context, that is, when the mind has to combine cues from the sound and the referent. Imagine, for example, that someone is uncertain whether they heard "pea" or "bee", does this uncertainty make them rely more on the referent (e.g., the object being pointed at)? Vice versa, if they are not sure if they saw a bee or a fly, does it make them rely more on the sound? More importantly, when input in both modalities is uncertain to varying degrees, do they weight each modality according to its relative reliability (which is the optimal strategy), or do they over-rely on a particular modality (which is a sub-optimal strategy)?

@@

First, in the case of audio-visual speech, modalities are understood to differ only in terms of their reliability. In a referential context, however, modalities also differ in terms of their roles in the referential process: the auditory input represents the symbol (i.e., the word) whereas the visual input represents the referent. It has been suggested that because of its referential property, speech is a privileged signal for humans, starting in infancy [see @vouloumanos2014 for a review].\footnote{There is a debate over whether speech is privileged for infant and adults for the same reasons. Whereas Waxman and colleagues suggest that speech is privileged for both infants and adults because of its ability to refer, Sloutsky and colleagues suggest that speech might not have a referential status from the start. Rather, according to Sloutsky and colleagues, speech seems to be prefered by infants only because of a low level auditory "overshadowing".} Thus, in a referential context, it is possible that we do not treat the auditory and visual modalities as equivalent sources of information. Instead, there could be a bias for the auditory modality beyond what is expected from informational reliability alone.

@@

We tested the ideal observer model's predictions in three behavioral experiments where we varied the source of uncertainty. In Experiment 1, audio-visual tokens were ambiguous with respect to their category membership only. In Experiment 2, we intervened by adding background noise to the auditory modality, and in Experiment 3, we intervened by adding background noise to the visual modality. In all experiments, participants were quantitatively near-optimal, though overall response precision was slightly lower than expected. Moreover, in Experiment 1 where neither of the modalities was perturbed with background noise, participants weighed auditory and visual cues according to the relative reliability predicted by the optimal model. In other words, we found no evidence for a modality bias towards either the auditory or the visual modality. However, in Experiment 2 and 3, participants over-relied on one modality when the other modality was perturbed with additional noise. <!-- In the discussion, we talk about the broad implications of these results and especially in relation to some developmental findings in early word learning [e.g., @stager1997]. -->

@@

In this section we describe the multimodal cue combination model. First, we briefly introduce the experimental paradigm, and second we explain how behavior in this paradigm can be characterized optimally with an ideal observer model.

@@

```{r task, fig.cap = "Overview of the task. In the audio-visual condition, participants are first presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test. The test may differ from the target in both the auditory and the visual components. After these two presentations, participants press `same' (i.e., the same category as the target) or `different' (not the same category). The auditory-only and visual-only conditions are similar to the audio-visual condition, except that only the sounds are heard, or only the pictures are shown, respectively.", fig.align = "center", out.width = "400px"}
knitr::include_graphics("pictures/task.png", dpi = 108)
```

This task is similar to the task introduced by @sloutsky2003, and used in subsequent studies in both children and adults [see @robinson2010 for a review]. However, in this original task, participants are asked whether or not the two audio-visual presentations are *identical*. Such task would allow us to probe audio-visual encoding, but not necessarily language understanding. The latter requires---in addition to the perceptual encoding---a categorical judgement, i.e, determining whether or not two similar tokens are members of the same phonological/semantic category. Thus, our task is category-based: Participants are asked to press "same" if they think the second item (the test) belonged to the same category as the first (target) (e.g.,  dog-/aba/), even if there is a slight difference in the sound, in the referent, or in both. They are instructed to press "different" only if they think that the second stimulus was an instance of the other word category (cat-/ada/). The task also includes trials where pictures are hidden (audio-only) or where sounds are muted (visual-only). These unimodal trials provide us with the participants' evaluation of the probabilistic information present in the auditory and visual categories. This unimodal distributions are used as inputs to the optimal combination model, described below.

@@

The ideal observer model combines probabilistic information from the auditory and visual modalities, and makes a categorical decision. How can this task be characterized in a formal way, and what would be the optimal behavior? If the stimuli were continuous (e.g., size, location), cues derived from each modality would be characterized with their sensory noise variance only, and (assuming conditional independence) their combined information would be characterized with a simple linear combination rule where the combined estimate is an average of the individual estimates, each weighted according to its relative reliability [@ernst02; @alais04]. However, in our case decision is made over categorical variables (phonemes and concepts), the optimal model should take into account, not only the noise variability around an individual perceptual estimate, but also its categorical variability (i.e., the probability with which the estimate belongs to a given category).

One way to account for the categorical variability is to model the decision step which maps sensory input onto task-dependent categories [@Bankieris17; @bejjanki2011]. We represent the decision step as follows (see Figure\ \@ref(fig:model) for illustration). If we assume that both the visual and the auditory cues vary along a single dimension, an audio-visual signal $(a,v)$ can be represented as a point in this two-dimensional space, and a word category can be characterized with a bivariate probability distribution over audio-visual tokens.\footnote{For the sake of clarity, we assume here that the sensory noise around this point is small/negligible compared to the categorical variability. But see below how the noise can be integrated when its effect is relatively large} Each category is peaked around the most typical audio-visual token, and the probability of membership in the category (the likelihood probability) decreases gradually as the auditory token deviates from the typical sound instance (depending on the variance of the auditory category) and/or as the visual token deviates from the typical visual instance (depending on the variance of the visual category). <!--To illustrate how categorical decision is made within this formal framework, suppose we receive an ambiguous image which can be either of a cat or of a dog, associated with a slightly noisy auditory input that sounds more like "cat". If the judgment is based on the picture only, this would lead to a random choice between cat and dog. However, if the judgement is based on both the picture and the sound (which is less ambiguous), the combined information should lead to a preference for the word cat  we illustrate the cateorical decision making. Audio-visual points along this line lead to random responses, meaning that both the auditory and the visual components are maximally ambiguous. Audio-visual points that lay on either side of the decision threshold lead to decisions in favor of one word or the other. This means that at least one component of this audio-visual token is not totally ambiguous, which leads to a diambiguated combined information. -->

A crucial feature of the ideal observer model is that it makes use of the entire probability distribution of a given category in order to make optimal decisions [e.g., @clayard08]. This means that the probability of picking a word (the posterior probability) should mirror closely the probability of the combined information (the likelihood probability). For instance, the model becomes less ambiguous and more certain as the tokens get closer to the typical instance of the category. In what follows, we formalize this sense of optimality using a Bayesian framework.
<!--characterized with a likelihood function, e.g., $P(S_V | S)$ and $P(S_A | S)$ for a visual and an auditory cue, respectively. If these cues are conditionally independent, their combined information can be written as $P(S_V, S_A | S)=P(S_A | S)P(S_V | S)$. The maximum likelihood estimate of this combined information $\hat{S}$ is an average of the individual maximum liklihoods $\hat{S}=w_V\hat{S_V}+w_A\hat{S_A}$, each wighted according to its relative reliability:
$w_V=\frac{1/\sigma_V^2}{1/\sigma_V^2 + 1/\sigma_A^2}$, and $w_A=\frac{1/\sigma_A^2}{1/\sigma_V^2 + 1/\sigma_A^2}$.

@@

An auditory category $A$ and a visual category $V$ are defined as distributions in the auditory and the visual dimensions, respectively. A word category $W$ is defined as a bivariate distribution in the audio-visual. More specifically, we define it as the joint distribution of the auditory and the visual categories.  For simplicity, we assume both auditory and visual categories to be normally distributed. That is, if $w=(a,v)$ is an audio-visual token, then:
$$ p(a | A) \sim  N(\mu_A, \sigma^2_A) $$
where $\mu_A$ and $\sigma^2_A$ are respectively the mean and the variance of the auditory category,
$$ p(v | V) \sim  N(\mu_V, \sigma^2_V) $$
where $\mu_V$ and $\sigma^2_V$ are  the mean and the variance of the visual category, and

@@

This assumption simply says that, given a word-object mapping, e.g., W=("cat"-CAT), variation in the way "cat" is pronounced does not correlate with changes in any visual property of the object CAT, which is a valid assumption.\footnote{Note that this assumptions is more adequate in the case of arbitrary associations such as ours, and less so in the case of redundant association such as audio-visual speech. In the latter, variation in the pronunciation is expected to correlate, at least to some extent, with lip movements.}

Now we turn to the crucial question of modeling how optimal decision should proceed based on the probabilistic information in the auditory and the visual modalities. In our task, we have two word categories: dog-/aba/ ($W_1$) and cat-/ada/ ($W_2$).\footnote{This mapping is randomized in the experiments.} Participants can be understood as choosing one of these two word categories (Figure\ \@ref(fig:model)). For an ideal observer, the probability of choosing category 2 when presented with an audio-visual instance $w=(a,v)$ is the posterior probability of this category:
$$
p(W_2 | w)=\frac{p(w|W_2)p(W_2)}{p(w|W_2)p(W_2)+p(w|W_1)p(W_1)}
$$
Using our assumption that the cues are uncorrelated, we have
$$p(w | W) = p(a,v| W) = p(a| A)p(v| V)$$
Under this assumption, the posterior probability reduces to:\footnote{See the appendix for the details of the derivation}
\begin{equation}
 p(W_2 | w)=\frac{1}{1+(1+\epsilon)\exp(\beta_0+\beta_aa+\beta_vv)}
\end{equation}
where 
$$1+\epsilon=\frac{p(W_1)}{p(W_2)}$$
$$\beta_0=\frac{\mu^2_{A2}-\mu^2_{A1}}{2\sigma^2_{A}}+\frac{\mu^2_{V2}-\mu^2_{V1}}{2\sigma^2_{V}}$$

@@

The parameter $\epsilon$ represents the differential between the categories' prior probabilities. However, since the identity of word categories is randomized across participants, $\epsilon$ measures, rather, a response bias to "same" if $\epsilon > 0$, and a response bias to "different" if $\epsilon < 0$. We expect a general bias towards answering "different" because of the categorical nature of our same-different task: When two items are ambiguous but perceptually different, participants might have a slight preference for "different" over "same".

@@

## Auditory and Visual models

Besides the optimal model, we also test the predictions of two baselines models: a visual model, which assumes that participants rely only on visual information, and an auditory model, which assumes that participants rely only on auditory information. If participants rely on both the auditory and the visual modalities, the optimal model would explain more variance in human responses than the visual model alone or the auditory model alone.

@@

A systematic comparison of the optimal and the descriptive models would allow us, not only to quantify how much people deviate from optimality, but also to understand precisely how they deviate from this optimality. Let $\sigma^2_{A}$ and $\sigma^2_{V}$ be the values of the variances used in the optimal model (derived from the unimodal conditions), and $\sigma^2_{Ab}$ and $\sigma^2_{Vb}$ be the values observed through the descriptive model in the bimodal condition. Deviation from optimality is measured in two ways. First, we measure the change in the values of the variances specific to each modality, that is, how $\sigma^2_{A}$ compares to $\sigma^2_{Ab}$, and how $\sigma^2_{V}$ compares to $\sigma^2_{Vb}$. Second, we measure changes in the proportion of the visual and auditory variances, i.e., how $\frac{\sigma^2_{A}}{\sigma^2_{V}}$ compares to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}$. The first measure allows us to test if precision changes for each modality when we move from the unimodal to the bimodal conditions. The second allows us to test whether or not the weighting scheme follows the prediction of the optimal model.  The reason we used the proportion of the variances as a measure of cross-modal weighing is because it corresponds to the slope\footnote{Or more precisely the absolute value of the slope} of the decision threshold in the audio-visual space (Figure\ \@ref(fig:model)). The decision threshold is defined as the set of values in this audio-visual space along which the posterior is equal to 0.5. Formally speaking, the decision threshold takes the following form:

@@

If the absolute value of the slope derived from the descriptive model is greater than that of the optimal model, the corresponding shift in the decision threshold indicates that participants have a preference for the auditory modality in the bimodal case. Similarly, a smaller absolute value of the slope would lead to a preference for the visual modality. The limit cases are when there is exclusive reliance on the auditory cue (a vertical line), and where there is exclusive reliance on the visual (a horizontal line).

It follows that there are three possible scenarios describing how humans can deviate from optimality. These scenarios are illustrated in Figure\ \@ref(fig:subOptim) and are as follows:

1) Both variances may increase, but their proportion remains the same. That is, $\sigma^2_{Ab} \geqslant \sigma^2_{A}$ and $\sigma^2_{Vb} \geqslant \sigma^2_{V}$, but  $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}} \approx \frac{\sigma^2_{A}}{\sigma^2_{V}}$. In this case, sub-optimality would be due to increased randomness in human responses in the bimodal condition. However, this randomness would not affect the relative weighting of both modalities, i.e., participants would still weigh modalities according to the relative reliability predicted by the optimal model.

2) The auditory variance increases at a higher rate.  That is, $\sigma^2_{Ab} \gg \sigma^2_{A}$ and $\sigma^2_{Vb} \geqslant \sigma^2_{V}$, leading to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}} > \frac{\sigma^2_{A}}{\sigma^2_{V}}$. In this case, sub-optimally would consist not only in participants being more random in the bimodal condition, but also in having a systematic preference for the visual modality, even after accounting for informational reliability.

3) The visual variance increases at a higher rate. That is, $\sigma^2_{Vb} \gg \sigma^2_{V}$, and  $\sigma^2_{Ab} \geqslant \sigma^2_{A}$, leading to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}} > \frac{\sigma^2_{A}}{\sigma^2_{V}}$. This case is the reverse of case 2, i.e., in addition to increased randomness in the bimodal condition, there is a systematic preference for the auditory modality, even after accounting for informational reliability.

```{r subOptim, fig.cap = "Illustration using simulated data showing the example of a prediction made by the optimal model (top), and the three possible ways human participants can deviate from this prediction (bottom). These cases are the following: 1) The variance increases equally for both modalities, but the weighting scheme (characterized by the decision threshold) is optimal, 2) The auditory variance increases at a higher rate, leading to a preference for the auditory modality, and 3) The visual variance increases at a higher rate, leading to a preference for the visual modality", fig.align = "center", out.width = "400px"}
knitr::include_graphics("pictures/sub-optimal", dpi = 108)
```

We tested the predictions of the optimal model in three experiments. In Experiment 1, we studied the case where bimodal uncertainty was due to ambiguity in terms of category membership, without any additional background noise. In Experiment 2 and 3 we added background noise on top of ambiguity in category membership.

@@

In this Experiment, we start with testing the predictions in the case where uncertainty is due to ambiguity in terms of category membership only, that is, according to the following weighting scheme:  $$\beta_a \propto \frac{1}{\sigma^2_{A}}$$ $$\beta_v \propto \frac{1}{\sigma^2_{V}}$$ We do not introduce any additional noise in the background, and we assume that sensory noise is negligible compared to categorical variability.

@@

```{r bimodal, fig.cap = "Human responses vs. models' predictions in the bimodal condition. Shape represents auditory distance from the target, and color represents visual distance from the target.", fig.align = "center", out.width = "400px", fig.width=7, fig.height=8}
